Reduced Nucleotide Diversity on a Plant Y Chromosome Following Recent Recombination Suppression

### 2016-08-16 manuscript.tex
minor edits

\begin{abstract}
X and Y chromosomes differ in effective population size ($N_{e}$), rates of recombination, and exposure to natural selection, all of which can significantly affect levels of genetic variability within the genome. On Y chromosomes with suppressed recombination, both positive and negative selection can remove genetically linked neutral variation, resulting in a reduction in $N_{e}$ and the level of polymorphism on Y compared to X chromosomes or autosomes. However, non-selective factors including biased sex ratios and high variance in male fitness can also reduce Y-linked $N_{e}$, making it difficult infer the causes of low Y-chromosome diversity. Here, we investigate the factors affecting X- and Y-linked polymorphism during the early stages of plant sex chromosome evolution in \textit{Rumex hastatulus} (Polygonaceae). Strikingly, we find that nucleotide diversity on the Y chromosome is 40 fold lower than on the X chromosome. We demonstrate that the magnitude of this reduction is not consistent with a purely neutral model based on the observed occurrence of female biased sex ratios in this species, or by reduced male $N_{e}$ caused by high variance in male fitness. Rather, using forward simulations, we show that the diversity reduction on the Y is consistent with the predicted effects of purifying selection removing deleterious mutations and linked neutral variation. Given the recent origin of \textit{R. hastatulus} sex chromosomes, our results indicate that the low level of sequence diversity commonly observed on ancient Y chromosomes might evolve soon after sex chromosomes originate, with background selection playing an important role during the early stages of their evolution.
\end{abstract}

@@

In contrast to the situation on the X chromosome, our data indicate a strong and consistent diversity reduction on the Y chromosome, and we now consider several models that might explain reduction.

@@

 A highly reduced level of Y chromosome variability is a predicted outcome of models Y-chromosome evolution based on the removal of deleterious mutations and linked neutral variants (e.g., background selection) (\X refs), ...and several cases....(\X refs), but the occurrence of female-biased sex ratios in this species has also been predicted to lower Y diversity in the absence of such selective interference. Moreover, such a reduction in Y-linked diversity could be accented due to high variance in male reproductive success (Figure 1). This is not unusual in annual plants such as \textit{R. hastatulus}, which commonly exhibit extensive phenotypic plasticity in plant size and flower production (\X Harper 1977). Conceivably, such plastic differences among plants, owing to spatial heterogeneity in resources, could influence the intensity of male-male competition, and thus affect variance in male reproductive success. Given that male plants in this wind-pollinated species produce large amounts of pollen, and that female flowers are uniovulate, strong competition among males is expected.

 In common with most flowering plants we do not have marker-based estimates of the variance in male reproductive success in \textit{R. hastatulus}. However, ... by modeling...we evaluated whether such an effect could result in a level of Y/A diversity similar to what we observed by comparing our results to predictions from a  model for the joint effects of sex ratio bias and male reproductive variance (see Methods). Estimates of sex ratio bias in \textit{R. hastatulus} have been estimated, and they are consistently female-biased, ranging from $N_{m}/(N_{m}+N_{f})=0.4$ to $N_{m}/(N_{m}+N_{f})=0.35$ (with a mean of 0.62; N=46 \citep{pickup2013influence}.The Y/A diversity ratio predicted from this sex ratio, assuming a Poisson distribution of offspring numbers, is approximately 0.2 (Table1, Figure 3). Our estimated mean Y/A ratio, however, is 0.02, which is significantly lower ($\textit{P}<0.0001$). Indeed, even for higher values for the variance in male reproductive success (\X), the corresponding predictions for Y/A are all significantly higher than our observed ratio (\X table). Finally, neutral models that could not be rejected, i.e., in which a highly female-biased sex ratio (\X) and a high variance in male reproductive success predicted an Y/A ratio that fit our data (\X), simultaneously predicts X/A ratios that are significantly different from what we observed (\X Figure). Thus, our results indicate that the combined effects of sex ratio bias and variance in reproductive success cannot jointly explain our observed levels of X and Y diversity in \textit{R. hastatulus}.

>>>>

\begin{abstract}
X and Y chromosomes differ in effective population size ($N_{e}$), rates of recombination, and exposure to natural selection, all of which can significantly affect levels of genetic variability within the genome. On Y chromosomes with suppressed recombination, both positive and negative selection can remove genetically linked neutral variation, resulting in a reduction in $N_{e}$ and the level of polymorphism on Y compared to X chromosomes or autosomes. However, non-selective factors including female biased sex ratios and high variance in male fitness can also reduce Y-linked $N_{e}$, making it difficult infer the causes of low Y-chromosome diversity. Here, we investigate the factors affecting X- and Y-linked polymorphism during the early stages of plant sex chromosome evolution in \textit{Rumex hastatulus} (Polygonaceae). Strikingly, we find that nucleotide diversity on the Y chromosome is 40 fold lower than on the X chromosome. We demonstrate that the magnitude of this reduction is not consistent with a purely neutral model based on the occurrence of female biased sex ratios in this species, or by reduced male $N_{e}$ caused by high variance in male fitness. Rather, using forward simulations, we show that the diversity reduction on the Y is consistent with the predicted effects of purifying selection removing deleterious mutations and linked neutral variation. Given the recent origin of \textit{R. hastatulus} sex chromosomes, our results indicate that the low level of sequence diversity commonly observed on ancient Y chromosomes can evolve soon after sex chromosomes originate, with background selection playing an important role during the early stages of their evolution.
\end{abstract}

@@

In contrast to the situation on the X chromosome, our data indicate a strong and consistent diversity reduction on the Y chromosome, and we now consider several models that might explain this reduction.

@@

 A highly reduced level of Y chromosome variability is a predicted outcome of models Y-chromosome evolution based on the removal of deleterious mutations and linked neutral variants (e.g., background selection), but the occurrence of female-biased sex ratios in this species has also been predicted to lower Y diversity in the absence of such selective interference (\X ref). Moreover, reduced Y-linked diversity could be accented by high variance in male reproductive success (Figure 1). This is not unusual in annual plants such as \textit{R. hastatulus}, which commonly exhibit extensive phenotypic plasticity in plant size and flower production (\X Harper 1977). Conceivably, such plastic differences among plants, owing to spatial heterogeneity in resources, could influence the intensity of male-male competition, and thus affect variance in male reproductive success. Given that male plants in this wind-pollinated species produce large amounts of pollen, and that female flowers are uniovulate, strong competition among males is expected.

 In common with most flowering plants we do not have marker-based estimates of the variance in male reproductive success in \textit{R. hastatulus}. However, by comparing our empirical estimates of diversity to predictions from models that jointly predict the effects on diversity of sex ratio bias and male reproductive variance, we evaluated whether these effects could explain the level of Y/A diversity that we observed (see Methods). Conditioning on estimates of sex ratio bias in \textit{R. hastatulus} that have been estimated, ranging from $N_{m}/(N_{m}+N_{f})=0.4$ to $N_{m}/(N_{m}+N_{f})=0.35$ \citep{pickup2013influence}, the predicted Y/A diversity ratio  under a Poisson distribution of offspring numbers, is approximately 0.2 (Table1, Figure 3). Our estimated mean Y/A ratio, however, is 0.02, which is significantly lower ($\textit{P}<0.0001$), suggesting that the sex ratio effect alone is insufficient to explain our data. For larger values for the variance in male reproductive success (\X), the corresponding predictions for Y/A are all significantly higher than our observed ratio (\X table). 
 Finally, neutral models that could not be rejected, i.e., in which a highly female-biased sex ratio (\X) and a high variance in male reproductive success predicted an Y/A ratio that fit our data (\X), simultaneously predicts X/A ratios that are significantly different from what we observed (\X Figure). Thus, our results indicate that the combined effects of sex ratio bias and variance in reproductive success cannot jointly explain our observed levels of X and Y diversity in \textit{R. hastatulus}.
